% chapters/11_pretraining.tex
% Part of: AI / LLM / Agents / Hardware Notes

\section{Pretraining Pipelines \& Data Curation}

\subsection{Web-scale data collection}

% TODO

\subsection{Deduplication \& quality filtering}

% TODO

\subsection{Data mixing strategies}

% TODO

\subsection{Pretraining objectives (CLM, MLM, etc.)}

\textbf{Causal Language Modelling (CLM).} The standard objective for autoregressive
models: predict the next token given all preceding tokens. Loss is cross-entropy over
the vocabulary at each position. Also called \emph{next-token prediction}. Used by
GPT, LLaMA, Mistral, and most production LLMs.

\textbf{Masked Language Modelling (MLM).} The BERT-family objective: randomly mask
$\approx$15\% of input tokens and predict them from bidirectional context. Produces
richer contextual representations but is not generative.

\textbf{Span corruption (T5).} Replace random contiguous spans with a sentinel token
and predict the replaced spans. Combines CLM and MLM; enables both encoding and
generation.

\textbf{The generality of CLM: beyond natural language.} The CLM objective is
domain-agnostic --- it requires only a discrete token sequence with learnable
structure. This has enabled foundation models trained on non-textual ``languages'':

\begin{itemize}
  \item \textbf{Code:} Codex, StarCoder, and DeepSeek-Coder treat source code as a
    token sequence, learning syntax, APIs, and idioms from next-token prediction.

  \item \textbf{Genomics (DNA/RNA):} EVO\,2 treats nucleotide sequences (A, T, G, C)
    as a four-letter alphabet and applies CLM at single-nucleotide resolution.
    The crucial insight: DNA from living organisms is an
    \emph{evolutionary self-labelled dataset}. Because organisms must function to
    survive and reproduce, every sequence in the training corpus has already been
    filtered by billions of years of natural selection. Lethal mutations are absent;
    conserved sequences are functionally critical. CLM thus receives an implicit
    supervision signal at a scale and quality that no human-annotated dataset could
    match. See Chapter~51.

  \item \textbf{Protein sequences:} ESMFold and ProtTrans apply masked language
    modelling to amino acid sequences. The evolutionary conservation signal is even
    stronger at the protein level --- most protein positions are under stronger
    selective pressure than individual nucleotides.

  \item \textbf{Music and audio tokens:} discretised audio (EnCodec, SoundStream)
    makes music generation a CLM task.

  \item \textbf{Agent trajectories:} CLM on (observation, action) sequences models
    agent behaviour offline --- the basis of Decision Transformer.
\end{itemize}

\begin{intuitionbox}
  The CLM objective does not care \emph{what} the tokens represent, only that the
  sequence has learnable structure. Any domain where the future is predictable from
  the past --- code, DNA, protein, audio, actions --- is in principle a CLM problem.
  The quality of implicit supervision depends on how strongly the domain's structure
  constrains valid sequences: evolutionary selection pressure makes genomic CLM
  arguably more information-dense per token than internet text.
\end{intuitionbox}

