% chapters/12_finetuning.tex
% Part of: AI / LLM / Agents / Hardware Notes

\section{Fine-tuning: SFT, LoRA, QLoRA}

\subsection{Supervised fine-tuning (SFT)}

\textbf{Supervised Fine-Tuning (SFT)} continues training a pretrained base model on a
curated dataset of (instruction, response) pairs. The objective is identical to
pretraining (cross-entropy over the response tokens), but the data distribution shifts
from raw internet text to human-authored demonstrations of the desired behaviour.

\textbf{What SFT accomplishes:}
\begin{itemize}
  \item \textbf{Instruction following:} the base model learns to respond to explicit
    task prompts rather than continuing text in the pretraining style.
  \item \textbf{Format compliance:} teaches the model to use specific response
    formats (markdown, JSON, conversational tone).
  \item \textbf{Surface alignment:} reduces obvious harmful, off-topic, or low-quality
    outputs, though it does not align values as robustly as RLHF.
\end{itemize}

\textbf{Data quality over quantity.} 1,000 carefully curated, diverse SFT examples
can outperform 100,000 noisy ones. The LIMA paper (Zhou et al.\ 2023) demonstrated
that 1K high-quality examples produced competitive instruction-following with models
trained on orders of magnitude more data. Data curation and deduplication are the
dominant effort in SFT pipelines.

\textbf{Masking.} During SFT training, loss is computed only on the response tokens;
the instruction/prompt tokens are masked from the loss computation. This prevents the
model from ``learning'' to output prompts.

\subsection{Parameter-efficient fine-tuning (PEFT)}

Full fine-tuning updates all model parameters, which costs as much GPU memory as
pretraining and risks catastrophic forgetting when data is limited.
\textbf{Parameter-Efficient Fine-Tuning (PEFT)} methods update only a small subset of
parameters while keeping most of the model frozen.

\textbf{Common PEFT families:}
\begin{itemize}
  \item \textbf{Adapter layers:} insert small bottleneck layers between transformer
    components; only adapter weights are trained ($<1\%$ of parameters).
  \item \textbf{Prefix tuning:} prepend trainable ``virtual tokens'' to the input;
    gradients flow only through these prefix vectors.
  \item \textbf{Prompt tuning:} even simpler --- only a small prefix embedding is
    trained; the model weights are entirely frozen.
  \item \textbf{LoRA} (see below): the dominant PEFT method for LLMs.
\end{itemize}

\textbf{Why PEFT works:} pretrained models learn rich representations; most task-
specific adaptation lies in a low-dimensional subspace of the weight update. PEFT
explicitly exploits this intrinsic dimensionality.

\subsection{LoRA \& AdaLoRA}

\begin{paperbox}[title={LoRA: Low-Rank Adaptation (Hu et al.\ 2022)}]
  Freezes the pretrained weight matrix $W_0 \in \mathbb{R}^{d \times k}$ and
  injects a trainable low-rank decomposition $\Delta W = BA$ where
  $B \in \mathbb{R}^{d \times r}$, $A \in \mathbb{R}^{r \times k}$, $r \ll \min(d,k)$.
  The modified forward pass is $h = W_0 x + \frac{\alpha}{r} B A x$.
  At inference, the LoRA weights can be merged ($W' = W_0 + \frac{\alpha}{r}BA$)
  so there is \emph{zero} additional inference latency.
\end{paperbox}

\textbf{Key properties:}
\begin{itemize}
  \item Typically applied to the $Q$, $K$, $V$, and output projection matrices;
    sometimes also to MLP layers.
  \item Rank $r = 4$--$64$ is common; smaller $r$ acts as stronger regularisation.
    $\alpha$ is a scaling hyperparameter (often set equal to $r$ so the factor is 1).
  \item Trainable parameters: $r \cdot (d + k)$ per matrix $\approx 0.1$--$1\%$ of
    total parameters for typical LLM sizes.
  \item $A$ is initialised with random Gaussian; $B$ is initialised to zero, so
    $\Delta W = 0$ at the start of training (no disruption to the pretrained model).
\end{itemize}

\textbf{AdaLoRA} adaptively allocates the rank budget across weight matrices based on
importance scores (derived from SVD of the gradient signal). High-importance matrices
receive more rank; low-importance ones are pruned. This avoids the uniform rank
assumption of vanilla LoRA.

\begin{intuitionbox}
  LoRA is the ``surgical strike'' fine-tuning method: instead of re-training the
  full weight matrix, it adds a small residual correction in a low-dimensional
  subspace. The pretrained knowledge is preserved; only the task-specific delta is
  learned.
\end{intuitionbox}

\subsection{QLoRA \& quantized fine-tuning}

\begin{paperbox}[title={QLoRA (Dettmers et al.\ 2023)}]
  Combines quantization with LoRA to make fine-tuning billion-parameter models
  feasible on a single consumer/research GPU.
\end{paperbox}

\textbf{Three technical innovations:}

\begin{itemize}
  \item \textbf{4-bit NormalFloat (NF4):} a quantization format specifically designed
    for normally-distributed weights (as neural network parameters are). Minimises
    quantization error for Gaussian distributions; performs better than INT4 or FP4.

  \item \textbf{Double quantization:} quantizes the quantization constants themselves
    (the per-block scaling factors) from FP32 to FP8, saving $\approx$0.5 bits per
    parameter on top of the NF4 savings.

  \item \textbf{Paged optimisers:} uses NVIDIA Unified Memory to offload optimiser
    states to CPU RAM when GPU memory is exhausted during gradient accumulation.
    Avoids OOM crashes during long forward passes with large batch sizes.
\end{itemize}

\textbf{Workflow:} (1) load base model in NF4 4-bit precision; (2) attach FP16/BF16
LoRA adapters to the frozen quantized base; (3) train only the LoRA adapters. The
base model is never de-quantized during training; dequantization happens on-the-fly
during each forward/backward pass.

\textbf{Practical impact.} QLoRA made it possible to fine-tune a 65B-parameter model
on a single 48\,GB A100 GPU (or a 13B model on a single 24\,GB consumer GPU), enabling
researchers and practitioners without large compute budgets to fine-tune frontier-scale
models.

\begin{notebox}
  QLoRA introduces a small accuracy penalty versus full-precision LoRA due to the 4-bit
  base model. For most task-specific fine-tunes the gap is negligible; for highly
  precise scientific/mathematical tasks it may matter.
\end{notebox}
