% chapters/16_quantization.tex
% Part of: AI / LLM / Agents / Hardware Notes

\section{Quantization \& Model Compression}

Quantization reduces the numerical precision of model weights (and optionally
activations) from 32-bit floating point to smaller representations. The goal is to
reduce memory footprint, memory bandwidth requirements, and sometimes compute cost,
with minimal impact on output quality.

\subsection{Post-training quantization (PTQ)}

\textbf{PTQ} quantizes a fully trained model without any re-training. It is fast and
requires only a small calibration dataset.

\textbf{Weight quantization.} Weights $W$ are mapped to discrete levels:
\[
  W_q = \text{round}\!\left(\frac{W}{\Delta}\right), \quad
  \Delta = \frac{\max(W) - \min(W)}{2^b - 1}
\]
where $b$ is the bit-width and $\Delta$ is the scale factor (learned per tensor,
channel, or group).

\textbf{Granularity.}
\begin{itemize}
  \item \textbf{Per-tensor:} single scale for entire weight matrix. Cheapest; most
    accuracy loss.
  \item \textbf{Per-channel:} separate scale per output channel. Better accuracy with
    modest overhead.
  \item \textbf{Per-group (block quantization):} scale per group of $g$ weights
    (e.g.\ $g=128$). Used in GPTQ and AWQ; good balance of accuracy and efficiency.
\end{itemize}

\textbf{Activation quantization.} Quantizing activations at runtime (as opposed to
just weights) is harder because activation distributions vary per input. Dynamic
quantization recomputes scales on-the-fly; static quantization uses calibration
statistics.

\subsection{Quantization-aware training (QAT)}

\textbf{QAT} simulates quantization noise during training so the model learns to be
robust to it. Weights and activations are quantized in the forward pass using
straight-through estimators (STE) for gradients. QAT consistently outperforms PTQ,
especially at very low bit-widths (INT4 and below), at the cost of requiring a
training run.

\textbf{FP8 training} (H100): NVIDIA Transformer Engine supports mixed-precision
training with FP8 (E4M3 or E5M2) for matrix multiplications. The master weights are
kept in BF16/FP32; FP8 is used for the compute pass. This gives roughly $2\times$
throughput vs FP16 on H100 tensor cores.

\subsection{GPTQ, AWQ, GGUF formats}

\textbf{GPTQ} (Frantar et al.\ 2023): second-order PTQ method. Uses the Hessian of
the weight quantization error to iteratively quantize weights while compensating for
errors using the remaining unquantized weights in the same row. Achieves high quality
at 4-bit weight quantization with a one-shot calibration pass. Standard for 4-bit LLM
deployment on GPU.

\textbf{AWQ} (Activation-aware Weight Quantization, Lin et al.\ 2024): observes that
not all weights are equally important --- weights corresponding to \emph{salient input
channels} (those with large activation magnitudes) should be quantized more carefully.
AWQ scales salient channels before quantization so their effective precision is
preserved. Faster calibration than GPTQ with competitive accuracy; widely used in
TGI and vLLM.

\textbf{GGUF} (llama.cpp): file format for storing quantized models for CPU/edge
inference. Supports a range from Q2\_K to Q8\_0 (2-bit to 8-bit), with mixed-precision
``K-quants'' that use higher precision for attention weights and lower for FFN weights.
The primary format for local inference with llama.cpp, Ollama, and LM Studio.

\begin{notebox}
  Common deployment rule of thumb: GPTQ/AWQ INT4 for GPU serving, GGUF Q4/Q5 for CPU
  inference. Both achieve perplexity within 1--3\% of the FP16 baseline on typical
  LLM sizes.
\end{notebox}

\subsection{Pruning \& distillation}

\textbf{Pruning} removes weights or entire structures (heads, layers) whose contribution
to the output is small.
\begin{itemize}
  \item \textbf{Unstructured pruning:} zeroes individual weights based on magnitude or
    gradient sensitivity. High compression ratios possible, but sparse weights require
    custom hardware/kernels to benefit at inference.
  \item \textbf{Structured pruning:} removes entire heads, channels, or layers. Output
    is a smaller dense model that runs efficiently on standard hardware. Used in DistilBERT
    (half the layers of BERT) and ShearedLLaMA.
\end{itemize}

\textbf{Knowledge distillation.} A smaller \emph{student} model is trained to mimic a
larger \emph{teacher} model's output distribution, intermediate representations, or
both. The student learns a compressed version of the teacher's knowledge.
\begin{itemize}
  \item \textbf{Response distillation:} student minimises KL divergence from the
    teacher's soft output probabilities (softer labels than hard ground-truth carry
    more information about the distribution).
  \item \textbf{Feature distillation:} student additionally matches intermediate
    layer activations of the teacher.
  \item \textbf{Data augmentation via distillation:} generate synthetic training data
    from the teacher to train the student (used in TinyLlama, Orca, etc.).
\end{itemize}

\begin{paperbox}[title={BitNet / 1-bit LLMs (Ma et al.\ 2024)}]
  Trains LLMs with weights constrained to $\{-1, 0, +1\}$ (BitNet\,b1.58). Each weight
  is represented in 1.58\,bits on average. Inference can replace matrix multiplications
  with addition/subtraction operations. At 3B+ parameters, BitNet\,b1.58 matches
  FP16 LLM quality while using significantly less memory and energy.
  Requires training from scratch (not a PTQ method); the extreme low-precision
  constraint must be baked into the training objective.
\end{paperbox}

\begin{warnbox}
  Distillation risks: the student inherits biases and errors from the teacher, including
  hallucinations. Benchmark overfitting is a concern when the teacher is used to generate
  training data. Legal gray area exists when commercial model outputs are used as
  distillation data without license clearance (see also Chapter~52).
\end{warnbox}
