% chapters/20_tool_use.tex
% Part of: AI / LLM / Agents / Hardware Notes

\section{Tool Use, Function Calling \& MCP}

\subsection{Function calling APIs}

Modern LLM APIs (OpenAI, Anthropic, Google) support \emph{tool use} (also called
function calling): the model can emit a structured tool-invocation request instead of
plain text, the caller executes it, and the result is fed back in the next turn.

\textbf{Anthropic Messages API structure}, as observed in Claude Code JSONL logs:

\begin{itemize}
  \item A \texttt{tool\_use} content block in an assistant message contains
    \texttt{id}, \texttt{name}, and \texttt{input} (a JSON object matching the tool's
    schema). Multiple tool calls can appear in the same content array.

  \item Tool results are returned by the caller as a \texttt{user} turn containing a
    \texttt{tool\_result} content block with the matching \texttt{tool\_use\_id}. This
    means the API's call/result cycle always alternates user/assistant turns, even though
    the ``user'' message is machine-generated.

  \item Each result in the JSONL carries a \texttt{toolUseResult} wrapper with
    \texttt{stdout}, \texttt{stderr}, \texttt{interrupted} (boolean, set if the user
    killed the process), \texttt{isImage} (for screenshot tools), and
    \texttt{noOutputExpected}.

  \item The \texttt{sourceToolAssistantUUID} field links the result back to the exact
    assistant event that issued the call, enabling causal tracing across the log.
\end{itemize}

\textbf{Tool loop termination.} The loop ends when the model emits a
\texttt{stop\_reason} of \texttt{end\_turn} (no more tool calls) or
\texttt{max\_tokens}. Agents must handle the \texttt{tool\_use} stop reason to detect
mid-response tool calls vs.\ a completed response.

\subsection{Model Context Protocol (MCP)}

MCP (Anthropic, 2024) is an open protocol that standardises how agents connect to
external tools and data sources. It separates the \emph{tool implementation} (an MCP
server) from the \emph{model} (an MCP client), so any compliant agent can use any
compliant server without custom glue code.

\begin{itemize}
  \item MCP servers expose \emph{tools}, \emph{resources} (data the model can read),
    and \emph{prompts} (reusable prompt fragments) over a JSON-RPC interface via
    stdio or HTTP/SSE.
  \item In Cowork, MCP connectors provide authenticated access to external services
    (Slack, Asana, Google Drive, etc.). The agent never handles raw credentials;
    authentication is managed by the MCP server layer.
  \item A central registry (\texttt{mcp-registry}) lets agents discover available
    connectors at runtime, enabling dynamic tool acquisition.
\end{itemize}

\subsection{Tool schema design}

A tool schema is the structured definition that tells the model what a tool does and
how to call it. Schema quality directly determines whether the model chooses tools
correctly and forms valid arguments.

\textbf{Schema fields (standard across APIs):}
\begin{itemize}
  \item \textbf{Name:} a unique, snake\_case identifier. The model uses this to
    reference the tool in its tool\_use output. Collision-free naming matters in
    large tool sets.
  \item \textbf{Description:} natural-language explanation of what the tool does,
    when to use it, and when \emph{not} to use it. This is the primary signal the
    model uses for tool selection --- a bad description causes wrong tool choice far
    more often than a bad parameter schema.
  \item \textbf{Parameters:} a JSON Schema object defining the tool's input fields,
    their types, descriptions, constraints (enum, pattern, minimum/maximum), and which
    are required vs.\ optional.
  \item \textbf{Tool call ID:} a unique tracking ID (assigned by the API) linking each
    invocation to its result in multi-tool conversations.
\end{itemize}

\textbf{Design heuristics:}
\begin{itemize}
  \item Keep tool names semantically distinct; \texttt{search\_web} vs.\
    \texttt{search\_docs} is clearer than \texttt{search1} vs.\ \texttt{search2}.
  \item Descriptions should include \emph{when to use} and \emph{when not to use}.
    Example: ``Use this tool to look up current weather. Do not use for historical
    weather data; use \texttt{weather\_history} instead.''
  \item Limit required parameters. Optional parameters reduce model friction; use
    sensible defaults and document them.
  \item For tools with large output surfaces (filesystem reads, web fetches), document
    the output format so the model can parse results correctly.
  \item Test schemas by inspecting which tool the model selects for edge-case prompts;
    iterate on descriptions before parameters.
\end{itemize}

\textbf{Combined with RAG:} tool-calling and RAG complement each other. The model
can use a \texttt{search} tool to retrieve documents, then consolidate multiple
retrieval results and reduce redundant API calls, producing a richer and more
contextually grounded response than either approach alone.

\subsection{Error handling in tool loops}

Tool failures must be communicated back to the model so it can adapt rather than
silently stall:

\begin{itemize}
  \item The Anthropic API's \texttt{tool\_result} block accepts an \texttt{is\_error:
    true} flag alongside the error message. The JSONL log confirms this: permission
    errors (e.g.\ \texttt{EACCES}) are returned as error results with the same
    \texttt{tool\_use\_id}, not as exceptions in the outer call.

  \item \textbf{Interrupted tools.} The \texttt{interrupted} field in
    \texttt{toolUseResult} signals that a long-running process was killed mid-execution
    (e.g.\ the user cancelled). The agent should treat partial output as unreliable.

  \item \textbf{Retry heuristics.} Well-designed agents distinguish between transient
    errors (network timeout --- retry), permission errors (wrong path --- try alternate
    approach), and hard failures (bad schema --- surface to user). Blindly retrying
    permission errors wastes turns and context budget.

  \item \textbf{Context-budget awareness.} Because tool results consume input tokens,
    agents should prefer tools that return concise outputs (e.g.\ \texttt{Grep} over
    full-file \texttt{Read}) when context is constrained.
\end{itemize}

