% chapters/23_planning.tex
% Part of: AI / LLM / Agents / Hardware Notes

\section{Planning: MCTS, CoT, o1-style Reasoning}

\subsection{Classical planning (STRIPS, PDDL)}

% TODO

\subsection{Monte Carlo Tree Search (MCTS)}

% TODO

\subsection{Chain-of-thought \& scratchpads}

Chain-of-thought prompting produces a step-by-step reasoning trace inside the model's
output (see Chapter~17 for prompting mechanics). In the context of planning, CoT
enables a model to \emph{decompose} a complex goal into sub-goals before committing
to an action.

\textbf{Scratchpad variants:}
\begin{itemize}
  \item \textbf{Visible CoT:} reasoning is part of the assistant turn and is seen by
    the user. Interpretable but verbose.
  \item \textbf{Extended thinking / thinking blocks:} reasoning in a hidden scratchpad
    that is signed but not shown to the user (Anthropic Claude o-style). Allows deeper
    multi-step reasoning without polluting the response.
  \item \textbf{Scratchpad in-weights:} test-time compute scaling (o1/o3 style) uses
    learned ``thinking tokens'' at inference; see Chapter~49.
\end{itemize}

\subsection{ReAct \& agent planning loops}

\begin{paperbox}[title=ReAct: Synergizing Reasoning and Acting (Yao et al.\, 2023)]
  ReAct interleaves chain-of-thought reasoning traces with concrete tool-use actions
  in a tight loop, grounding abstract reasoning in real-world observations.
\end{paperbox}

\textbf{The ReAct loop} alternates three step types in sequence:

\begin{enumerate}
  \item \textbf{Thought} --- the model reasons about its current state, what
    information it lacks, and what action to take next (CoT-style natural language).
  \item \textbf{Action} --- the model invokes a tool (search, calculator, API call,
    code execution) with structured arguments.
  \item \textbf{Observation} --- the tool result is returned and appended to context;
    the model reads it and produces the next Thought.
\end{enumerate}

The loop terminates when the model emits a \textbf{Finish} action with a final
answer, when a maximum iteration count is reached, or when a confidence threshold
is satisfied.

\begin{center}
\begin{tikzpicture}[
  node distance=2.2cm,
  box/.style={rectangle, rounded corners=4pt, draw=aiblue, fill=infobg,
              minimum width=2.4cm, minimum height=0.9cm, align=center,
              font=\small\bfseries},
  finish/.style={rectangle, rounded corners=4pt, draw=aiteal, fill=okgreen,
                 minimum width=2.4cm, minimum height=0.9cm, align=center,
                 font=\small\bfseries},
  arr/.style={-{Stealth[length=6pt]}, thick, color=aigray}
]
  \node[box]   (thought)   {Thought};
  \node[box, right=of thought]  (action)    {Action\\(Tool Call)};
  \node[box, right=of action]   (obs)       {Observation\\(Tool Result)};
  \node[finish, below=1.2cm of action] (finish) {Finish};

  \draw[arr] (thought) -- (action);
  \draw[arr] (action)  -- (obs);
  \draw[arr] (obs.north) -- ++(0,0.7) -| (thought.north)
    node[midway, above, font=\scriptsize, color=aigray] {iterate};
  \draw[arr, dashed] (action.south) -- (finish)
    node[midway, right, font=\scriptsize, color=aiteal] {final answer};
\end{tikzpicture}
\end{center}

\textbf{Advantages over pure CoT:}
\begin{itemize}
  \item Reduces hallucination by grounding each reasoning step in retrieved evidence.
  \item Enables the model to course-correct mid-loop when an observation contradicts
    its prior reasoning.
  \item Produces an interpretable audit trail: thought → action → observation
    triples are logged (see JSONL session format, Chapter~52).
\end{itemize}

\textbf{Multi-agent ReAct:} a large orchestrator model handles high-level reasoning
and Thought/Finish steps while delegating Action steps (tool calls) to smaller
specialised subagents.

\textbf{ReWOO (Reason Without Observation):}
An alternative to ReAct that separates the planning and execution phases. Rather than
interleaving Thought and Observation, ReWOO generates a complete plan upfront (a
sequence of tool calls with dependencies), then executes them, then synthesises the
results. Benefits: fewer LLM calls (cheaper), plan is inspectable before execution.
Cost: cannot course-correct mid-plan based on intermediate observations.

\subsection{Plan-and-Execute}

Plan-and-Execute separates agent reasoning into two distinct phases, unlike ReAct
which interleaves them:

\begin{enumerate}
  \item \textbf{Planner phase:} the model receives the goal and produces a structured
    multi-step plan \emph{before any tool is called}. Example: ``1.~gather company
    background, 2.~retrieve financial data, 3.~compare competitors, 4.~write report.''
  \item \textbf{Executor phase:} a (potentially different, cheaper) model or module
    carries out each step in sequence, passing results forward.
\end{enumerate}

\textbf{Compared to ReAct:}

\begin{center}
\begin{tabular}{lll}
\toprule
 & \textbf{ReAct} & \textbf{Plan-and-Execute} \\
\midrule
Planning & Interleaved with actions & Upfront, then separate \\
Adaptability & High (course-corrects mid-loop) & Low (plan commits early) \\
LLM calls & More (each step) & Fewer for planning \\
Debuggability & Harder (emergent path) & Easier (plan is inspectable) \\
Best for & Open-ended, adaptive tasks & Structured, predictable tasks \\
\bottomrule
\end{tabular}
\end{center}

\textbf{ReWOO} is a close relative: both front-load planning; ReWOO additionally
chains tool-call dependencies explicitly in the plan so the executor knows which
earlier results to substitute (avoiding redundant re-retrieval).

\textbf{Hybrid approaches:} production agents often combine patterns --- an
orchestrator produces a plan (Plan-and-Execute style), then each step executes
using a ReAct inner loop that handles step-level surprises, with human-in-the-loop
approval at major plan transitions (LangGraph interrupt nodes).

\subsection{o1 / o3 test-time compute scaling}

% TODO

