% chapters/33_distributed_training.tex
% Part of: AI / LLM / Agents / Hardware Notes

\section{Distributed Training}

Modern large-scale LLM training combines multiple parallelism strategies simultaneously.
The NVIDIA Megatron team coined the term \textbf{5D Parallelism} to describe the five
orthogonal dimensions along which a training job can be distributed: Data (DP), Tensor
(TP), Pipeline (PP), Sequence/Context (SP/CP), and Expert (EP).

\begin{notebox}
  Each parallelism dimension targets a different bottleneck. Real training runs combine
  all five: a typical 70B job might use DP$\times$8, TP$\times$4, PP$\times$4, with
  SP inside TP regions, and EP added for MoE variants.
\end{notebox}

\subsection{Data parallelism (DDP)}

\textbf{Data parallelism} replicates the full model on every GPU and splits the global
mini-batch across GPUs. Each GPU computes a full forward and backward pass on its shard,
then gradients are synchronised with an all-reduce before the optimiser step.

\textbf{DDP (DistributedDataParallel)} in PyTorch buckets gradients and overlaps
all-reduce communication with the tail of the backward pass to hide latency.

\textbf{Fully Sharded Data Parallel (FSDP)} (see \S\,\ref{sub:fsdp}) extends DDP by
sharding optimiser states, gradients, \emph{and} parameters across DP ranks.

\subsection{Tensor parallelism}

\textbf{Tensor Parallelism (TP)} splits individual weight matrices across GPUs.
For a linear layer $Y = XW$ with $W \in \mathbb{R}^{d \times k}$:

\begin{itemize}
  \item \textbf{Column-parallel:} $W$ is split column-wise into $[W_1 | W_2]$ across
    $N$ GPUs. Each GPU computes $XW_i$ and the outputs are concatenated via all-gather.
  \item \textbf{Row-parallel:} $W$ is split row-wise; the input $X$ must be split
    correspondingly, and partial sums are reduced across GPUs with all-reduce.
\end{itemize}

Megatron-LM implements column-parallel and row-parallel linear layers with a single
all-reduce per transformer layer in each direction. Because all-reduce must complete
before the next layer, TP requires extremely low-latency interconnects (NVLink within
a node; TP is almost never used across nodes).

\subsection{Pipeline parallelism}

\textbf{Pipeline Parallelism (PP)} assigns contiguous groups of transformer layers to
different GPUs (``stages''). A micro-batch flows through each stage in sequence, then
the next micro-batch fills the pipeline. The \textbf{pipeline bubble} --- idle time
when a stage is waiting for upstream activations --- is the main inefficiency.

\textbf{1F1B schedule} (Megatron-LM): interleaves forward and backward micro-batches
across stages. Each stage alternates one forward pass with one backward pass, reducing
peak pipeline bubble from $O(p)$ to $O(1)$ relative to the number of stages $p$.

\begin{paperbox}[title={DualPipe (DeepSeek-V3, 2025)}]
  Bidirectional pipeline where micro-batches travel \emph{both} forward and backward
  through the pipeline simultaneously. Computation and communication are fully
  overlapped, reducing bubble fraction to near zero. Used in DeepSeek-V3 training
  alongside EP for MoE layers.
\end{paperbox}

\subsection{FSDP \& ZeRO optimizer}\label{sub:fsdp}

A single GPU cannot hold the full state of a large model. The \textbf{ZeRO} optimizer
(Zero Redundancy Optimizer, Microsoft DeepSpeed) eliminates the redundant copies of
state that vanilla DP creates.

\begin{center}
\begin{tabular}{lll}
\toprule
\textbf{Stage} & \textbf{What is sharded} & \textbf{Memory saving (vs.\ DP)} \\
\midrule
ZeRO-1 & Optimiser states & $4\times$ \\
ZeRO-2 & + Gradients & $8\times$ \\
ZeRO-3 & + Parameters & $64\times$ (linear in world size) \\
\bottomrule
\end{tabular}
\end{center}

PyTorch's \textbf{FSDP} implements ZeRO-3 natively: during the forward pass each GPU
all-gathers the full parameter shard, uses it, then immediately discards it. Gradients
are accumulated locally, then reduce-scattered. Only $1/N$ of the parameters live on
each GPU at rest.

\textbf{ZeRO-Infinity} (DeepSpeed) extends sharding to CPU RAM and NVMe storage,
enabling trillion-parameter training at the cost of higher latency for offloaded data.

\textbf{Frameworks:}
\begin{itemize}
  \item \textbf{Megatron-LM} (NVIDIA): reference implementation of TP + PP with
    1F1B scheduling; used to train GPT-3, Llama, Mistral, and most NVIDIA-scale models.
  \item \textbf{DeepSpeed} (Microsoft): ZeRO optimizer, activation checkpointing,
    offloading to CPU/NVMe. Often combined with Megatron via \emph{Megatron-DeepSpeed}.
  \item \textbf{Transformer Engine} (NVIDIA): mixed-precision training with FP8 on H100
    hardware; integrates directly into Megatron-LM and NeMo.
  \item \textbf{Megatron Bridge}: adapter to use Megatron-LM parallelism with other
    training stacks.
\end{itemize}

\subsection{3D parallelism \& Sequence/Context parallelism}

\textbf{3D parallelism} combines DP + TP + PP and is the standard recipe for hundred-
billion-scale training. Each node group runs TP internally (NVLink); PP connects node
groups sequentially; DP replicates the entire pipeline.

\textbf{Sequence Parallelism (SP)} (Megatron-LM): splits the sequence dimension along
the portions of a layer (LayerNorm, Dropout) that lie outside the TP-parallelised
matrix multiplications. This removes the remaining activation memory bottleneck in TP.

\textbf{Context Parallelism (CP)}: for very long contexts (100K+ tokens), the sequence
dimension is split across GPUs even during the attention computation itself. Ring
Attention and similar algorithms enable cross-GPU attention by passing keys and values
in a ring pattern, so each GPU sees the full sequence for attention while storing only
a shard of the activations.

\textbf{Expert Parallelism (EP)}: in Mixture-of-Experts (MoE) models each expert is
assigned to a different GPU. Token routing via all-to-all communication dispatches each
token to its selected expert and gathers the results back. EP combined with TP/PP/DP
gives \emph{5D parallelism}. See also: Mixtral of Experts, DeepSeek-MoE.

\begin{intuitionbox}
  Think of 5D parallelism as slicing a 5-dimensional tensor (model $\times$ data $\times$
  layers $\times$ sequence $\times$ experts) across a cluster of GPUs. Each dimension
  targets a different memory or compute bottleneck.
\end{intuitionbox}
