% chapters/44_interpretability.tex
% Part of: AI / LLM / Agents / Hardware Notes

\section{Interpretability \& Mechanistic Analysis}

Mechanistic interpretability (MI) aims to reverse-engineer neural networks into human-
understandable algorithms. Rather than treating a model as a black box, MI tries to
identify the \emph{circuits}, \emph{features}, and \emph{representations} the model
has learned and explain \emph{why} it produces particular outputs.

\subsection{Probing classifiers}

A \textbf{probing classifier} trains a simple linear (or shallow MLP) classifier on top
of a frozen model's internal activations to test whether a particular concept is linearly
encoded at a given layer.

\begin{itemize}
  \item High probe accuracy $\Rightarrow$ the concept is linearly represented at that
    layer. Low accuracy $\Rightarrow$ either the concept is not encoded, or it is encoded
    non-linearly.
  \item \textbf{Limitation:} a probe showing accuracy does not prove the model
    \emph{uses} that representation for its downstream behaviour. Probing is a
    necessary but not sufficient condition for functional relevance.
  \item Common probes: part-of-speech tags on intermediate layers, world-state probes
    in game-playing models (Othello-GPT), truth probes on factual knowledge.
\end{itemize}

\subsection{Superposition hypothesis}

Neural networks represent far more features than they have neurons. The
\textbf{superposition hypothesis} (Elhage et al.\ 2022, Anthropic) explains how:

\begin{itemize}
  \item Features (interpretable properties of the world) are encoded as \emph{directions
    in activation space}, not as individual neurons.
  \item The same neuron participates in encoding multiple features simultaneously
    (\textbf{polysemanticity}): a single neuron may respond to ``banana'',
    ``yellow'', \emph{and} ``the word `the'\,''.
  \item By spreading features across nearly-orthogonal directions in a high-dimensional
    space, the network can represent $N \gg d$ features in a $d$-dimensional activation
    space. The cost is interference between co-activating features.
  \item Features with lower frequency and higher importance get cleaner, more
    dedicated representations; rare but important features can avoid superposition.
\end{itemize}

\begin{intuitionbox}
  Superposition is the neural-network analogue of compressed sensing: you can recover
  $N$ sparse signals from $d < N$ measurements if the signals are rarely co-active.
  The network exploits the sparsity of natural concept co-occurrence.
\end{intuitionbox}

\subsection{Sparse autoencoders (SAE)}

Sparse autoencoders are the primary tool for \emph{decomposing} superposed
representations into interpretable monosemantic features.

\textbf{Architecture.} An SAE has an encoder that maps an activation vector
$\mathbf{x} \in \mathbb{R}^d$ to a sparse latent $\mathbf{f} \in \mathbb{R}^m$ ($m \gg d$),
and a decoder that reconstructs $\hat{\mathbf{x}}$:
\[
  \mathbf{f} = \text{ReLU}(W_e \mathbf{x} + \mathbf{b}_e), \quad
  \hat{\mathbf{x}} = W_d \mathbf{f} + \mathbf{b}_d
\]
Trained with an $L_2$ reconstruction loss plus an $L_1$ sparsity penalty on $\mathbf{f}$.
The $L_1$ penalty forces each activation to be explained by only a few features.

\textbf{Interpretation.} Each dimension of $\mathbf{f}$ corresponds to a ``feature'';
the decoder column $W_d[:, i]$ is that feature's direction in activation space.
If a feature's activating inputs cluster around a semantically coherent concept (e.g.\
``Python function definitions'', ``references to DNA''), it is labelled as that concept.

\begin{paperbox}[title={Towards Monosemanticity (Cunningham et al.\ 2023, Anthropic)}]
  Applied SAEs to a one-layer transformer. Found thousands of interpretable features
  including curve detectors, high-frequency features, and multimodal neurons. Validated
  that SAE features are causally relevant: steering on a feature's direction induces
  the corresponding behaviour.
\end{paperbox}

\begin{paperbox}[title={Scaling Monosemanticity (Templeton et al.\ 2024, Anthropic)}]
  Scaled SAEs to Claude\,3 Sonnet (a large production model) and extracted
  $\sim$34\,million features. Found high-level abstract concepts including features for
  the ``Assistant'' token that encode intent and agency. Demonstrated that feature
  steering on safety-relevant features (e.g.\ ``deception'') produces consistent
  behavioural changes.
\end{paperbox}

\textbf{Tooling.}
\begin{itemize}
  \item \textbf{SAELens}: open-source library for training, analysing, and evaluating
    SAEs across model families. Provides standardised training pipelines and
    interpretability metrics.
  \item \textbf{Neuronpedia}: web platform for browsing trained SAE features, viewing
    max-activating dataset examples, and testing feature steering.
\end{itemize}

\subsection{Circuit-level analysis}

A \textbf{circuit} is a minimal subgraph of model components (attention heads, MLP
neurons, residual stream paths) that is sufficient to explain a particular model
behaviour.

\textbf{Methodology.}
\begin{enumerate}
  \item Identify a \emph{task} (e.g.\ indirect object identification: ``John told
    Mary that \underline{\hspace{1em}} liked ice cream'' $\to$ ``she'').
  \item Use \textbf{activation patching} (see below) to identify which components
    causally matter for the task.
  \item Trace the information flow through the residual stream to characterise what
    each component contributes.
\end{enumerate}

\textbf{Canonical circuits found in transformers:}
\begin{itemize}
  \item \textbf{Induction heads} (Olsson et al.\ 2022): two-layer circuit that copies
    patterns from earlier context (\texttt{[A][B]\ldots[A]} $\to$ predict \texttt{B}).
    Mechanistically implement in-context learning.
  \item \textbf{Indirect Object Identification (IOI)} (Wang et al.\ 2022): a 26-head
    circuit in GPT-2 medium that correctly identifies indirect objects in sentences like
    ``When John and Mary went to the store, John gave a book to\ldots''
\end{itemize}

\textbf{Circuit-Tracer} is an automated tool for discovering circuits without manual
intervention; it uses gradient-based attribution to identify the most impactful paths
through the computation graph.

\subsection{Activation patching}

\textbf{Activation patching} (also called \emph{causal tracing}) is an intervention
technique that identifies which activations causally mediate a model behaviour.

\textbf{Protocol:}
\begin{enumerate}
  \item Run the model on a \emph{clean} input (correct behaviour) and a \emph{corrupted}
    input (incorrect behaviour, e.g.\ name swap). Record all activations.
  \item Run the corrupted input again, but \emph{patch} the activation at a specific
    layer and token position by substituting the clean activation.
  \item Measure how much the model's output recovers toward the clean behaviour.
    A large recovery means that activation is causally important.
\end{enumerate}

\begin{paperbox}[title={ROME / MEMIT (Meng et al.\ 2022)}]
  Used causal tracing on GPT-2 and GPT-J to localise factual associations to specific
  MLP layers in the middle of the network. Used this insight to design
  Rank-One Model Editing (ROME), which directly overwrites factual associations
  in those MLP weight matrices without fine-tuning.
\end{paperbox}

\textbf{Path patching} extends the technique by patching not just a single activation
but a specific \emph{path} between two components, enabling more precise circuit
decomposition.

\begin{warnbox}
  Activation patching can identify correlation between a component and a behaviour,
  but determining the full mechanism still requires understanding \emph{what} the
  component computes, not just \emph{that} it matters. Patching is the localisation
  step; circuit analysis is the explanation step.
\end{warnbox}
