% chapters/45_eval_methodology.tex
% Part of: AI / LLM / Agents / Hardware Notes

\section{Evaluation Methodology \& Benchmark Design}

\subsection{Evaluation taxonomy}

Evaluations exist on two orthogonal axes: \emph{what} is being measured and \emph{how}
it is measured.

\textbf{What is being measured:}
\begin{itemize}
  \item \textbf{LLM output quality:} fluency, coherence, factual accuracy of a single
    response. Standard metrics: BLEU, ROUGE, BERTScore, perplexity.
  \item \textbf{Agent task success:} did the agent complete the goal? Fundamentally
    different from output quality --- a response can be fluent but wrong; a task can
    succeed via a suboptimal path. Metrics: task completion rate, step accuracy,
    tool-call efficiency, latency, cost per run.
  \item \textbf{Safety and alignment:} does the model/agent refuse correctly, avoid
    harmful outputs, resist injection, and respect privacy constraints?
  \item \textbf{Robustness:} does performance hold under paraphrasing, adversarial
    inputs, or distribution shift?
  \item \textbf{Efficiency:} latency, token count, cost per query or task.
\end{itemize}

\textbf{How it is measured:}
\begin{itemize}
  \item \textbf{Exact match:} strict string equality. Fast and unambiguous; brittle for
    open-ended tasks.
  \item \textbf{Functional correctness:} test-based evaluation --- does the output
    pass unit tests? Standard for code evals (SWE-bench, Aider Polyglot, HumanEval).
  \item \textbf{Rubric scoring:} a fixed rubric grades outputs on specific dimensions.
    Can be applied by human raters or an LLM judge.
  \item \textbf{LLM-as-judge:} a stronger model rates the output against a rubric.
    Scalable; inherits the judge model's biases and inconsistencies.
  \item \textbf{Human rating:} annotators assess quality. Expensive but sets the gold
    standard for nuanced, open-ended tasks.
  \item \textbf{End-to-end environment:} agent runs in a sandboxed environment and
    outcomes are verified programmatically (WebArena, OSWorld, SWE-bench).
\end{itemize}

\begin{intuitionbox}[title=Key distinction: output quality vs task success]
  A model can generate a fluent, coherent answer that fails the actual task. A model
  can succeed at the task via an ugly, verbose path. For agents, always prefer
  task-success metrics over output-quality proxies.
\end{intuitionbox}

\subsection{Contamination \& data leakage}

\textbf{Benchmark contamination} occurs when examples from an evaluation benchmark
appear (exactly or near-exactly) in the model's training data. This inflates measured
performance beyond the model's true generalisation capability.

\textbf{Why it happens:}
\begin{itemize}
  \item Web-scale pretraining datasets inevitably contain benchmark text published
    online, in papers, or in tutorials.
  \item Synthetic data pipelines (where a stronger model generates training examples)
    can inadvertently reproduce benchmark questions when the teacher model has seen them.
  \item Fine-tuning on large scraped corpora (GitHub, Stack Overflow, coding forums)
    can pull in code that is identical to benchmark test cases.
\end{itemize}

\textbf{Why it is hard to catch:}
\begin{itemize}
  \item Near-duplicates (paraphrased, variable-renamed) are harder to detect than
    exact copies and can still give unfair advantage.
  \item Developers may not check for contamination until after a model ``beats''
    a benchmark --- at which point the incentive to look is low.
  \item Some contamination is partial: a model may have seen the problem statement
    but not the ground-truth answer, giving a weaker but real advantage.
\end{itemize}

\textbf{Detection methods:}
\begin{itemize}
  \item \textbf{N-gram overlap:} check training data for $n$-gram matches with
    benchmark items. Computationally cheap; misses paraphrases.
  \item \textbf{Perplexity probing:} the model assigns lower perplexity to text it
    has memorised. Benchmark items with unusually low perplexity are suspect.
  \item \textbf{Canary strings:} embed unique strings in private benchmarks; if they
    appear in training data, contamination is confirmed.
  \item \textbf{Held-out splits:} maintain a private benchmark partition never
    published online.
\end{itemize}

\begin{warnbox}[title=Real-world example: PewDiePie's coding model]
  After training a coding model and achieving a score that appeared to beat ChatGPT's
  ChatGPT-4 on Aider Polyglot, the author discovered small amounts of contamination
  in his training data. He discarded the benchmark result and retrained on clean data
  --- demonstrating that contamination is easy to miss even in small, manually curated
  datasets. The final clean score (39.1\%) was ultimately higher, but the contaminated
  result was invalid regardless of the magnitude of overlap.
\end{warnbox}

\subsection{Human vs automated evaluation}

\begin{center}
\begin{tabular}{lll}
\toprule
 & \textbf{Automated} & \textbf{Human} \\
\midrule
Speed & Fast (seconds--minutes) & Slow (hours--days) \\
Cost & Cheap & Expensive \\
Scale & Unlimited & Limited by annotator pool \\
Nuance & Low--medium & High \\
Consistency & High (deterministic) & Variable (inter-rater variance) \\
Coverage & Easy to run at every PR & Typically periodic \\
Best for & Regression testing, CI & Gold standard, edge cases \\
\bottomrule
\end{tabular}
\end{center}

\textbf{Automated eval strategies:}
\begin{itemize}
  \item \textbf{Unit-test / functional:} execute generated code and check test passage.
    High fidelity for coding tasks; requires a sandboxed execution environment.
  \item \textbf{Regex / exact match:} good for factual Q\&A with known answers.
  \item \textbf{LLM-as-judge:} a stronger model scores outputs on a rubric.
    Practical at scale; biased toward its own output style and sensitive to prompt
    phrasing. Should be validated against human judgements.
  \item \textbf{Reference-free metrics:} BERTScore, METEOR measure semantic
    similarity to a reference; useful but imperfect for open-ended generation.
\end{itemize}

\textbf{Human eval strategies:}
\begin{itemize}
  \item \textbf{Pairwise comparison:} raters choose which of two outputs is better.
    Easier and more reliable than absolute ratings; enables Elo-style leaderboards
    (e.g.\ LMSYS Chatbot Arena).
  \item \textbf{Rubric-anchored rating:} raters score on a scale (1--5) against
    explicit criteria. Inter-rater agreement must be measured (Cohen's $\kappa$).
  \item \textbf{Task completion:} rater attempts to complete a task using the model's
    output. Produces a binary or partial-credit score.
\end{itemize}

\textbf{Regression testing workflow:}
When a model, prompt, or retrieval config changes, run the full automated eval suite
before and after and compare distributions. Flag regressions in specific categories
(coding, reasoning, safety) before deploying. Human review is reserved for cases
where automated metrics disagree or a novel failure mode is suspected.

\subsection{Statistical significance in evals}

\textbf{Benchmark scores are noisy.} LLMs are non-deterministic at temperature $> 0$,
many benchmarks use sampling to select a subset of problems, and small dataset sizes
amplify variance. A single reported number is rarely sufficient.

\textbf{Real-world variance observed:} On the Aider Polyglot coding benchmark, scores
for the same model checkpoint varied across runs --- 14\%, 15\%, and 16\% from
consecutive runs, and top-line frontier model trackers show week-to-week swings of
5--10 percentage points on identical benchmark versions. This is not model degradation;
it is inherent stochasticity.

\textbf{Best practices:}
\begin{itemize}
  \item \textbf{Report distributions, not points.} Run $n \geq 5$ trials; report mean
    and standard deviation. ``$25 \pm 2\%$'' is more honest than ``25\%''.
  \item \textbf{Statistical tests.} For comparing two model versions, use a paired
    $t$-test or bootstrap confidence interval on per-example scores. A 1-point
    improvement is rarely significant on a 100-item benchmark.
  \item \textbf{Enough items.} A 50-item benchmark has very wide confidence intervals.
    $n \geq 500$ items is a reasonable floor for reliable comparisons.
  \item \textbf{Fixed temperature / seed.} For reproducibility, fix temperature to 0
    (greedy decoding) or fix the random seed. Report which was used.
  \item \textbf{Never cherry-pick runs.} Running the benchmark 10 times and reporting
    only the best result (``god run'') is a form of data dredging --- the expected
    maximum of 10 samples is substantially higher than the true mean.
\end{itemize}

\begin{warnbox}[title=The ``god run'' anti-pattern]
  Running a benchmark repeatedly and reporting the highest score is statistically
  invalid. If the true mean performance is 25\% and you run 10 trials, the expected
  maximum will be several points higher simply by chance. Always report mean $\pm$ std
  across a fixed number of trials decided in advance.
\end{warnbox}

