% chapters/47_agent_evals.tex
% Part of: AI / LLM / Agents / Hardware Notes

\section{Evals for Agents \& Agentic Tasks}

\subsection{WebArena, OSWorld}

% TODO — benchmark-specific details to be added when covered

\subsection{AgentOps \& observability}

\textbf{AgentOps} (agent operations) is the emerging discipline of managing the full
lifecycle of autonomous agents in production, analogous to DevOps for software and
MLOps for models.

\textbf{Why agents are harder to operate than models:}
\begin{itemize}
  \item Non-deterministic: same input can produce different action sequences.
  \item Autonomous: they initiate actions, chain tasks, and make decisions without
    per-step human approval.
  \item Multi-turn: errors compound across steps; a wrong action in step 3 may only
    surface in step 10.
  \item Multi-system: agents call external APIs, databases, and sub-agents; failure
    can originate anywhere in the chain.
\end{itemize}

\textbf{AgentOps lifecycle phases:}
\begin{enumerate}
  \item \textbf{Development:} specify objectives and constraints, map task
    dependencies, define data pipelines and expected tool schemas.
  \item \textbf{Testing:} evaluate agent performance in a sandbox environment with
    simulated tool responses before production exposure.
  \item \textbf{Monitoring:} observe results at session, trace, and span levels. Track
    actions taken, API calls made, latency, total duration, and LLM cost per run.
  \item \textbf{Feedback:} log agent mistakes and surface them for review; mechanisms
    for triggering a corrected re-run or updating instructions.
  \item \textbf{Governance:} guardrails and policies constraining agent behaviour;
    regulatory compliance (GDPR, HIPAA); audit trails for accountability.
\end{enumerate}

\textbf{IBM / watsonx observability stack} (OpenTelemetry-based):
\begin{itemize}
  \item Built on OpenTelemetry (OTEL) standards for vendor-neutral instrumentation.
  \item Automatic and manual instrumentation across major frameworks (LangChain,
    CrewAI, AutoGen, watsonx Orchestrate, OpenAI Agents SDK).
  \item \textbf{Session replays:} step-by-step observation of an agent run, including
    every tool call, intermediate result, and reasoning step.
  \item \textbf{Multi-trace workflow views:} see how a task flows across multiple
    sessions, agents, and tool calls in a single dashboard.
  \item \textbf{Cost analysis:} trace LLM token spend across providers (Azure, AWS,
    IBM) per session or per task type.
  \item \textbf{Trajectory exploration:} inspect which paths through the agent's
    action space led to success vs.\ failure.
  \item AI-powered analytics layer for iterative benchmarking toward optimisation.
\end{itemize}

\textbf{LangSmith} (LangChain's eval and observability platform):
\begin{itemize}
  \item Automatically traces all model and tool calls to spot errors and latency
    outliers.
  \item Supports testing performance under different model configurations without
    code changes.
  \item Not limited to LangChain; integrates with any LLM application.
  \item \textbf{LLM-as-a-judge evaluators:} use a stronger model to score agent
    outputs on correctness, helpfulness, and safety criteria.
  \item Aims to bridge the gap between prototype performance and production-quality
    reliability.
\end{itemize}

\textbf{AgentOps ecosystem} (17+ tools as of 2024):
Agenta, LangSmith, TruLens, AgentOps.ai, Weights \& Biases Weave, Arize Phoenix,
Braintrust, Helicone, and others. Most focus on trace capture, cost tracking, and
evaluation scoring.

\subsection{SWE-bench Verified}

% TODO

\subsection{Tool-use evals}

Evaluating tool-using agents is fundamentally different from evaluating static
question-answering models:

\begin{itemize}
  \item \textbf{Trajectory correctness:} does the agent reach the right answer
    \emph{via the right sequence of tool calls}? A correct answer via a lucky wrong
    path is not a reliable success.
  \item \textbf{Tool selection accuracy:} does the agent choose the appropriate
    tool for each sub-task, and avoid calling tools unnecessarily?
  \item \textbf{Argument quality:} are the arguments passed to tools well-formed,
    correctly typed, and semantically appropriate?
  \item \textbf{Error recovery:} when a tool returns an error, does the agent
    diagnose and adapt, or loop in confusion?
  \item \textbf{Efficiency:} how many tool calls does the agent make vs.\ the
    minimum required? Excess calls inflate cost and latency.
\end{itemize}

\subsection{Long-horizon task evaluation}

Long-horizon tasks span many steps, multiple tools, and potentially multiple sessions.
Evaluation is harder because:

\begin{itemize}
  \item \textbf{Partial credit:} an agent that completes 8 of 10 required steps is
    meaningfully better than one that completes 2, but binary pass/fail metrics miss
    this. Step-level scoring is preferred.
  \item \textbf{State coupling:} later steps depend on earlier ones; an error early
    in the trajectory can make all subsequent steps fail even if the agent's logic
    is otherwise sound.
  \item \textbf{Non-determinism:} re-running the same task may produce different
    action sequences. Eval suites should run multiple trials and report distributions,
    not single results.
  \item \textbf{Sandboxed environments:} long-horizon evals (WebArena, OSWorld,
    SWE-bench) provide controlled environments with real or simulated backends so
    agents can take real actions without affecting production systems.
  \item \textbf{Human-in-the-loop baselines:} comparing agent task completion rate
    and time-to-completion against human workers provides the most meaningful
    benchmarks for enterprise use cases.
\end{itemize}

\subsection{Safety evals for agents}

Agent safety evaluation goes beyond behavioural alignment to cover the agent's
interaction with real-world resources:

\begin{itemize}
  \item \textbf{Prompt injection resistance:} does the agent correctly refuse to
    follow instructions embedded in tool results, web pages, or documents? Measured
    by injecting adversarial instructions into observed content and checking whether
    the agent acts on them.
  \item \textbf{Minimal-footprint compliance:} does the agent request only the
    permissions it needs, prefer reversible over irreversible actions, and avoid
    storing sensitive data unnecessarily?
  \item \textbf{Refusal accuracy:} does the agent correctly refuse prohibited actions
    (permanent deletion, credential entry, sharing permissions) while not over-refusing
    legitimate requests?
  \item \textbf{Escalation / confirmation behaviour:} does the agent pause and request
    human approval at the right points (novel situations, irreversible actions, high
    stakes) rather than proceeding autonomously?
  \item \textbf{Data leakage:} does the agent avoid routing sensitive context to
    unapproved model backends or external endpoints? (See also OpenShell privacy router,
    Chapter~36.)
\end{itemize}

\subsection{Agent failure modes}

Understanding failure modes drives the most impactful eval improvements.

\begin{itemize}
  \item \textbf{Hallucination:} the agent asserts facts not supported by evidence.
    Mitigations: RAG grounding, citation requirements, fact-check subagent,
    ``unknown'' fallback when evidence is absent.

  \item \textbf{Looping:} the agent repeats a tool call or reasoning cycle without
    progress. Mitigations: max-step limit, repeated-state detection, explicit stop
    conditions, or upfront planning (ReWOO).

  \item \textbf{Tool misuse:} wrong tool, malformed arguments, or over-calling.
    Mitigations: strong tool descriptions (Chapter~20), schema validation, explicit
    routing rules.

  \item \textbf{Context drift:} the agent loses track of the original goal across
    many steps. Mitigations: persistent task state, goal re-injection, summarised
    intermediate outputs.

  \item \textbf{Over-delegation:} spawning too many subagents for a task that could
    be handled directly, inflating cost and latency. Mitigation: delegate only when
    specialisation gives a measurable quality win; budget subagents.

  \item \textbf{Sycophancy:} telling the user what they want to hear rather than
    the accurate answer. Particularly dangerous in report-generation and evaluation
    agents. Measured by checking whether the agent contradicts demonstrably false
    user premises.
\end{itemize}

