% chapters/48_ssm_mamba.tex
% Part of: AI / LLM / Agents / Hardware Notes

\section{State Space Models: Mamba \& RWKV}

State Space Models (SSMs) are an alternative sequence-modelling paradigm to
transformers. They draw on classical control theory: a continuous-time linear dynamical
system maps an input signal $u(t)$ to an output $y(t)$ via a latent state $h(t)$.
Discretised for sequence data, SSMs offer $O(n)$ inference complexity with a constant-
size recurrent state, in contrast to the $O(n^2)$ attention matrix of transformers.

\subsection{S4 \& structured state spaces}

\begin{paperbox}[title={S4 (Gu et al.\ 2022)}]
  Structured State Space Sequence model. Parameterises the state transition matrix $A$
  using the \textbf{HiPPO} (High-order Polynomial Projection Operators) initialisation,
  which provably compresses the history of an input signal into orthogonal polynomial
  coefficients. This gives S4 the capacity to model extremely long-range dependencies
  that RNNs and transformers typically miss.
\end{paperbox}

\textbf{Continuous system.} The core equations are:
\[
  h'(t) = A h(t) + B u(t), \quad y(t) = C h(t) + D u(t)
\]
Discretised with step size $\Delta$:
\[
  h_k = \bar{A} h_{k-1} + \bar{B} u_k, \quad y_k = C h_k
\]
where $\bar{A} = e^{\Delta A}$ and $\bar{B} = (e^{\Delta A} - I) A^{-1} B$.

\textbf{Efficient training via convolution.} With fixed $A$, $B$, $C$, the recurrence
can be unrolled into a convolution $y = u * K$ where the convolutional kernel
$K = (CB, C\bar{A}B, C\bar{A}^2 B, \ldots)$ can be computed in $O(n \log n)$ with FFT.
This enables fast, parallelisable training while retaining $O(1)$ inference via the
recurrence.

\subsection{Mamba \& selective SSMs}

S4's state transition matrices $A$, $B$, $C$ are fixed (input-independent), giving a
linear time-invariant (LTI) system. Mamba breaks this assumption.

\begin{paperbox}[title={Mamba (Gu \& Dao 2023)}]
  Introduces a \textbf{selection mechanism}: the SSM parameters $B$, $C$, and the
  discretisation step $\Delta$ become functions of the input, making the model
  input-dependent (non-LTI). This allows the model to \emph{selectively} propagate
  or reset information based on content, analogous to the gating in LSTMs but
  with superior long-range capacity.
\end{paperbox}

\textbf{Key innovation.} Because the matrices are now input-dependent, the convolution
trick no longer applies. Mamba instead implements a hardware-aware \emph{selective scan}
algorithm that fuses the recurrence into a single GPU kernel operating in SRAM, achieving
efficient parallel computation during training while preserving $O(1)$ recurrent inference.

\textbf{Mamba architecture.} Each Mamba block is a gated MLP variant:
(1) the input is projected up; (2) one branch runs through the selective SSM;
(3) the other branch gates the SSM output; (4) the result is projected back down.
This mirrors the gated feedforward structure of modern LLMs.

\textbf{Mamba-2} simplifies the selective SSM with a multi-head structure (SSD:
Structured State Space Duality), establishing formal connections between SSMs and
linear attention. It is faster in practice with similar quality.

\subsection{RWKV: RNN-transformer hybrid}

\textbf{RWKV} (Peng et al.\ 2023) reframes transformer attention as a linear RNN using
a time-mixing mechanism inspired by both attention and classic RNNs. The key, value,
and receptance (gate) are computed as in a transformer, but attention is computed as a
cumulative weighted sum without the quadratic softmax matrix:
\[
  y_t = \frac{\sum_{\tau \leq t} e^{-(t-\tau) w + k_\tau} v_\tau}{\sum_{\tau \leq t} e^{-(t-\tau)w + k_\tau}}
\]
where $w$ is a learned decay parameter. This has $O(1)$ recurrent inference (like an
RNN) while being parallelisable during training.

\subsection{SSM vs attention trade-offs}

\begin{center}
\begin{tabular}{lll}
\toprule
\textbf{Property} & \textbf{Transformer (attention)} & \textbf{SSM / Mamba} \\
\midrule
Training complexity & $O(n^2)$ & $O(n)$ \\
Inference state & KV cache (grows with $n$) & Fixed recurrent state $O(1)$ \\
Long-context memory & Full context in theory & Fixed capacity hidden state \\
Positional inductive bias & Positional encodings needed & Baked in via state dynamics \\
Content addressability & Arbitrary key-value lookup & Approximate; recency bias \\
Hardware efficiency (A100) & FlashAttention helps & Selective scan; good SRAM use \\
\bottomrule
\end{tabular}
\end{center}

\textbf{When SSMs win:} audio, genomics, time series, and any domain where sequences
are extremely long and local patterns matter more than global content retrieval.

\textbf{When attention wins:} language reasoning, complex instruction following, tasks
requiring recall of specific content from many steps ago. The fixed-capacity SSM state
can ``forget'' distant information even if it is relevant.

\textbf{Hybrid models} (e.g.\ Jamba, Zamba) interleave Mamba and attention layers,
aiming to capture the memory efficiency of SSMs for most layers while retaining the
associative recall strength of attention in a few layers. This is an active research
frontier.

\begin{intuitionbox}
  Attention is like a database: you can look up any past token by content at any step,
  but the lookup table (KV cache) grows with context. An SSM is like RAM with a
  fixed size: you can only hold so much, and old information may be overwritten.
  Hybrid models give you both a large RAM and a lookup table, used selectively.
\end{intuitionbox}
