% chapters/49_test_time_compute.tex
% Part of: AI / LLM / Agents / Hardware Notes

\section{Test-time Compute \& Reasoning Models}

The standard scaling law says: more parameters $+$ more training compute $=$ better models.
A complementary axis is \textbf{test-time compute scaling}: investing more computation
at \emph{inference} time to improve answer quality on hard problems, without changing
the model weights.

\subsection{Scaling inference compute}

\textbf{The core idea.} A model with moderate training compute but extended test-time
thinking can outperform a much larger model on reasoning tasks. This is especially
powerful for problems with verifiable answers (math, code, formal proofs) where
correctness can be checked without human labelling.

\textbf{Mechanisms.}
\begin{itemize}
  \item \textbf{Serial compute:} generate more tokens before committing to an answer
    (thinking tokens, scratchpads). Allows the model to self-correct and explore
    reasoning branches.
  \item \textbf{Parallel compute:} sample many candidate solutions and select the
    best one (best-of-$N$, self-consistency). Parallelisable but $N\times$ more
    expensive.
  \item \textbf{Tree search:} combine parallel and serial compute by maintaining a
    tree of partial solutions evaluated by a process reward model (PRM). More
    computationally intensive but more directed than random sampling.
\end{itemize}

\textbf{Compute-optimal inference.} Snell et al.\ (2024) showed that for a fixed
inference budget, adaptive test-time compute (deciding \emph{how much} to think based
on problem difficulty) outperforms using the same budget uniformly.

\subsection{OpenAI o1 / o3 architecture (known)}

The \textbf{o1} family (OpenAI, 2024) represents the first widely-deployed test-time
compute scaling system. Limited public information suggests:

\begin{itemize}
  \item The model is trained with RL to produce extended ``reasoning traces'' (thinking
    tokens) before the final answer. The reasoning trace is hidden from users in the
    public API.
  \item Training uses a form of RLHF/GRPO where the RL reward signal is verifiable
    correctness (math benchmarks, code tests) rather than human preference.
  \item The model learns to use the thinking budget dynamically: easy problems produce
    short traces; hard problems produce long, branching internal reasoning.
  \item Performance scales predictably with thinking token budget: more thinking tokens
    $\Rightarrow$ higher accuracy on competition math (AIME, AMC) and coding benchmarks.
\end{itemize}

\textbf{DeepSeek-R1} (2025): an open-weight model trained with pure RL (GRPO) on
verifiable math/code problems. Demonstrates that the reasoning capability is learnable
without human-labelled thinking traces: the model discovers effective reasoning patterns
by reward signal alone. This is analogous to AlphaGo learning to play Go without human
game transcripts.

\subsection{Process reward models (PRMs)}

\textbf{Outcome Reward Models (ORMs)} give a reward only for the final answer. They are
simple to train but give sparse signal that does not help with long reasoning chains.

\textbf{Process Reward Models (PRMs)} give step-level rewards, evaluating the
correctness of each reasoning step rather than only the final conclusion. Benefits:
\begin{itemize}
  \item Denser training signal: the model receives feedback on every step of a
    multi-step proof or solution.
  \item Better tree search guidance: PRMs can score intermediate nodes in a search
    tree, enabling more efficient MCTS or beam search over reasoning paths.
  \item Interpretability: step-level rewards identify exactly where reasoning fails.
\end{itemize}

\textbf{Training PRMs.} Requires annotating not just final answers but individual
reasoning steps as correct/incorrect --- a more expensive labelling process. Math-
Shepherd (2024) and ``Let's Verify Step-by-Step'' (Lightman et al.\ 2023, OpenAI) are
key PRM papers.

\subsection{Best-of-N \& self-consistency}

\textbf{Best-of-$N$} (rejection sampling): sample $N$ independent completions from the
model and return the one with the highest reward model score. Simple and parallelisable;
quality improves roughly as $\log N$ with diminishing returns.

\textbf{Self-consistency} (Wang et al.\ 2022): for problems with a discrete final
answer, sample multiple reasoning chains and return the answer by majority vote.
Does not require a reward model; leverages the model's own distribution. Effective for
math (multiple paths often reach the same correct answer) but less useful for open-ended
tasks.

\textbf{Beam search with PRM:} maintain the top-$B$ partial solutions at each step,
scored by a PRM. More directed than random sampling; finds higher-quality solutions
with fewer total tokens generated.

\subsection{Thinking tokens \& scratchpads}

\textbf{Thinking tokens} are tokens generated by the model that are not part of the
visible output but are consumed as working memory before the final answer. Claude
implements this via \emph{extended thinking} blocks (see Chapter~22); OpenAI o1 uses
hidden reasoning traces.

\textbf{Chain-of-thought} is the prompt-level equivalent: eliciting step-by-step
reasoning as visible output text (see Chapter~17). Thinking tokens extend this by
making the scratchpad implicit and training the model end-to-end to use it efficiently.

\begin{paperbox}[title={Ouro: Scaling Latent Reasoning via Looped Language Models (2025)}]
  Trains a transformer to \emph{loop} its own hidden states as input for multiple
  iterations before generating output, similar to how recurrent networks process
  a sequence. Each loop is an implicit reasoning step operating in the latent
  space rather than the token space. The model learns to allocate more loops to
  harder problems. Because reasoning happens in the continuous latent space (not
  discrete tokens), the ``thinking'' is not interpretable to humans but can be
  computationally more efficient. This is an early instance of \emph{latent
  test-time compute scaling}.
\end{paperbox}

\begin{notebox}
  Ouro is conceptually close to the ``Repeat Your Self'' (RYS) approach and other
  looped/recurrent transformer variants (see also Chapter~48 on SSMs). The common
  thread: rather than expressing reasoning as token strings, these models perform
  computation in the latent space, trading interpretability for efficiency.
\end{notebox}
