% chapters/53_agentic_platforms.tex
% Part of: AI / LLM / Agents / Hardware Notes

\section{Agentic Platforms: Manus, Computer Use \& Beyond}

\subsection{Manus AI overview (Butterfly Effect)}

% TODO

\subsection{NVIDIA NemoClaw / OpenShell}

NemoClaw is NVIDIA's packaged enterprise distribution for agentic deployment, combining
three layers into a single installable stack (see Chapter~36 for full detail):

\begin{itemize}
  \item \textbf{OpenClaw} --- the agent/app layer (planning, tool use, file I/O).
  \item \textbf{OpenShell} --- the out-of-process runtime governor: sandbox +
    policy engine + privacy router. Sits between the agent and the real OS/network.
  \item \textbf{Nemotron} --- NVIDIA's open model family (reasoning, coding, safety,
    agentic) used as the local inference option when cloud routing is disallowed.
\end{itemize}

The distinguishing architectural idea is \emph{out-of-process policy enforcement}:
safety controls are applied at the OS level (Landlock, seccomp, network namespaces)
rather than relying on soft in-context instructions. Enterprise policy rules
(data residency, domain whitelists, filesystem scoping) are enforced by OpenShell
before any action executes, regardless of what the agent reasons it should do.

\textbf{OpenClaw as ``Agentic OS.''} Jensen Huang framed OpenClaw at GTC as the
operating system for \emph{personal agents} in the same way Windows was the OS for
personal computers: a universal runtime that abstracts hardware and exposes a stable
interface for applications. The analogy maps naturally --- OpenClaw manages resources
(tools, file systems, models), schedules work (cron jobs, task decomposition), spawns
sub-agents (analogous to processes), and handles I/O in any modality. Just as every
company needed a Linux strategy, an HTTP strategy, and a Kubernetes strategy, the
argument is that every software company now needs an \emph{OpenClaw strategy}.

\begin{notebox}[title={Nemotron Coalition}]
  NVIDIA announced a coalition of companies committing to build on and contribute to
  Nemotron\,4: \textbf{Black Forest Labs} (image generation), \textbf{Cursor}
  (AI coding IDE), \textbf{LangChain} (agent frameworks), \textbf{Mistral},
  \textbf{Perplexity}, \textbf{Reflection}, \textbf{Sarv} (India), \textbf{Thinking
  Machines}, and \textbf{Mirror/Moradi's lab}. The coalition accelerates the ecosystem
  of fine-tuned and specialised derivatives around a shared open base model.
\end{notebox}

\textbf{NVIDIA Open Model Families.} Beyond Nemotron (language), NVIDIA maintains six
frontier open model families spanning radically different domains:

\begin{center}
\renewcommand{\arraystretch}{1.3}
\begin{tabular}{lp{8cm}}
\toprule
\textbf{Model family} & \textbf{Domain} \\
\midrule
\textbf{Nemotron} & Language: reasoning, visual understanding, RAG, safety, speech \\
\textbf{Cosmos} & Physical AI: world generation and understanding for robotics/simulation \\
\textbf{Alpamo} & Autonomous vehicles: the world's first thinking/reasoning AV model \\
\textbf{Groot} & General-purpose humanoid and articulated robotics \\
\textbf{Bioneo} & Biology, chemistry, and molecular design \\
\textbf{Earth\,2} & Weather and climate forecasting via AI physics \\
\bottomrule
\end{tabular}
\end{center}

\begin{intuitionbox}
  These six families reflect the fact that no single model can serve every industry.
  Language intelligence (Nemotron) and physical-world intelligence (Cosmos, Alpamo,
  Groot) require fundamentally different training distributions, evaluation criteria,
  and safety considerations. NVIDIA's strategy: maintain frontier open models in each
  vertical so that industry partners can fine-tune domain-specific sovereign models on
  top.
\end{intuitionbox}

\subsection{SaaS \texorpdfstring{$\to$}{→} AaaS: the agentic transformation}

The enterprise software industry is undergoing a structural shift from
\textbf{Software-as-a-Service (SaaS)} to \textbf{Agentic-as-a-Service (AaaS)}.
Traditional SaaS companies build tools that \emph{humans} use: dashboards, workflow
engines, systems of record. Post-OpenClaw, those same companies are rebuilding their
products as agentic services --- autonomous agents that \emph{act} on a company's behalf,
not tools that humans operate.

\textbf{Enterprise IT before/after agentic systems.}
\begin{itemize}
  \item \textbf{Before (SaaS):} data centres hold structured business data; software
    exposes tools; humans (or digital workers) operate those tools; GSI consultants
    integrate everything. Value is created by humans wielding software.
  \item \textbf{After (AaaS):} agents sit on top of the same data stores and SaaS
    APIs, but they operate autonomously. The company's domain expertise is encoded in
    agent configurations rather than in UI workflows.
\end{itemize}

\textbf{Token budgets as a new engineering resource.} Huang predicted that every
engineer will eventually carry an \emph{annual token budget} --- analogous to a compute
or travel budget --- allowing them to amplify their work via AI inference. A software
engineer paid \$200K might receive a token budget worth \$100K+ on top, with the
expectation of 10$\times$ productivity. Token budget size is already emerging as a
recruiting differentiator in Silicon Valley.

\begin{notebox}
  The implication is that companies will simultaneously be \emph{token consumers}
  (buying compute to amplify their own engineers) and \emph{token manufacturers}
  (packaging their domain expertise as agentic services that customers pay per-token
  to access). AI factories (GPU clusters) become the production infrastructure for
  both sides of this equation.
\end{notebox}

\subsection{Anthropic Computer Use}

Anthropic's \emph{Computer Use} capability (released October 2024) exposes the model's
ability to observe screenshots and emit low-level desktop actions (mouse clicks, keyboard
input, scrolling). It is offered as a set of tool schemas (\texttt{computer},
\texttt{text\_editor}, \texttt{bash}) that the API caller is responsible for executing,
making the operator the ``hands'' that actually drive the OS.

\textbf{Cowork} is Anthropic's desktop-native product built on this foundation. Rather
than raw computer use, Cowork couples a sandboxed Linux VM with a rich tool layer
(filesystem access, web browsing, scheduled tasks, MCP connectors) to produce a
practical desktop agent for non-developers. Key architectural points observed directly:

\begin{itemize}
  \item The agent runs inside a per-conversation Ubuntu 22 VM spun up on the user's
    machine. Each session receives a randomly generated slug
    (e.g.\ \texttt{festive-nice-ramanujan}) used as both the VM directory name and the
    project identifier in the \texttt{.claude/} state store.
  \item A mounted workspace path (\texttt{/sessions/\{slug\}/mnt/\{user-folder\}})
    bridges the VM to the user's actual filesystem, while the VM's own root remains
    ephemeral and resets between tasks.
  \item A \textbf{Skills system} provides Claude with pre-written \texttt{SKILL.md}
    instruction files (e.g.\ \texttt{pptx}, \texttt{docx}, \texttt{pdf}, \texttt{xlsx})
    that encode best practices for creating specific document types, reducing trial-and-error
    at inference time.
  \item MCP (Model Context Protocol) connectors allow authenticated integration with
    external services (Slack, Asana, Google Drive, etc.) without the agent handling
    raw credentials.
\end{itemize}

\subsection{OpenAI Operator \& Agents SDK}

% TODO

\subsection{Browser-use \& web automation}

% TODO

\subsection{Desktop agent architectures}

Desktop agents must bridge the gap between an LLM's token-level reasoning and the
pixel/filesystem/network reality of a personal computer. A few recurring structural
patterns:

\begin{itemize}
  \item \textbf{Tool-first vs.\ screenshot-first.} Tool-first agents (Claude Code, Cowork)
    prefer structured tool calls (read file, run bash, fetch URL) over raw screenshot
    observation. Screenshot-first agents (OpenAI Operator, Manus) parse the visual DOM
    and emit pointer events. Tool-first agents are faster and cheaper for file/code tasks;
    screenshot-first generalizes to arbitrary GUI applications.

  \item \textbf{Workspace isolation.} A dedicated working directory (or VM) separates
    agent-generated artifacts from the user's files, preventing accidental overwrites.
    Cowork keeps a \texttt{/sessions/\{slug\}/} scratch space distinct from the mounted
    user folder.

  \item \textbf{State subdirectories.} The \texttt{.claude/} layout illustrates how
    desktop agents partition state: \texttt{projects/} for conversation logs,
    \texttt{plans/} for multi-step plans, \texttt{memory/} for persistent notes,
    \texttt{shell-snapshots/} for working-directory continuity across Bash calls, and
    \texttt{backups/} for rollback support.

  \item \textbf{Prompt caching at scale.} Long-running desktop sessions accumulate large
    system prompts. Tiered prompt caching (e.g.\ 5-minute and 1-hour TTL tiers in
    Anthropic's API) amortises the cost of re-encoding the same instructions on every
    turn.
\end{itemize}

\subsection{Security \& sandboxing for autonomous agents}

Autonomous agents operating on real filesystems and networks introduce a surface area for
both accidental harm and adversarial \emph{prompt injection}.

\textbf{Sandboxing layers} used in practice:
\begin{itemize}
  \item \textbf{VM isolation} --- each session runs in a separate lightweight VM so a
    runaway command cannot reach the host OS or other sessions.
  \item \textbf{Filesystem scoping} --- the agent's write access is limited to its own
    working directory; the user's files are mounted read-write only through an explicit
    workspace mount.
  \item \textbf{Permission tiers} --- Claude Code's \texttt{permissionMode} field
    (\texttt{default / full / restricted}) controls whether the agent can run arbitrary
    shell commands, write files, or is restricted to read-only actions.
  \item \textbf{Prohibited action lists} --- product-level policies (e.g.\ Cowork's
    security rules) enumerate actions the agent must refuse regardless of instruction:
    permanent deletion, modifying access controls, entering financial credentials, creating
    accounts.
\end{itemize}

\textbf{Prompt injection} is the primary adversarial threat: malicious instructions
embedded in observed content (web pages, emails, filenames, tool results) attempt to
hijack the agent's action stream. Defences include:
\begin{itemize}
  \item Treating all content from tool results as \emph{untrusted data}, not as
    instructions --- regardless of claimed authority (``admin override'', ``Anthropic
    staff'').
  \item Requiring explicit out-of-band user confirmation (in the chat interface) before
    acting on any instruction found inside a function result.
  \item Signing extended-thinking blocks (as in Claude's \texttt{signature} field) to
    detect tampering between turns in a long tool-use loop.
\end{itemize}

\subsection{Meta's agent research (LLaMA, CICERO)}

% TODO

